---
title: "Deaths from Overdoses"
author: "Bryana A.S. Thieret"
date: "2024-03-02"
format: 
  html: 
    page-layout: full
    messages: false
    echo: false
    warning: false
    fig-align: center
categories: [news]
---




```{r}
library(readxl)
library(tidyverse)
library(gridExtra)
library(here)

library(readxl)

od2 <- read_excel(here("posts/BlogPost3/overdose2.xlsx"), skip = 5)

#od2 <- read_excel("overdose2.xlsx", skip = 5) 
od2 = od2 |> slice(1:103) |> rename("Category" = ...1) |> mutate(gender = ifelse(
  Category == "Female" | Category == "Male", Category, NA
))

totalsODs = od2 |> pivot_longer(c(2:25), names_to = "year", values_to = "deaths") |>
  filter(is.na(gender)) |> 
  select(-gender)

list = c("Total Overdose Deaths", 
  "Any Opioid1", 
  "Prescription Opioids2",
  "Prescription Opioids AND Synthetic Opioids other than Methadone",
  "Prescription Opioids WITHOUT Synthetic Opioids other than Methadone",
  "Synthetic Opioids other than Methadone (primarily fentanyl)3",
  "Heroin4", 
  "Heroin AND Synthetic Opioids other than Methadone",
  "Heroin WITHOUT Synthetic Opioids other than Methadone",
  "Stimulants5a", 
  "Stimulants AND Any Opioid", 
  "Stimulants WITHOUT Any Opioid",
  "Stimulants AND Synthetic Opioids other than Methadone",
  "Stimulants WITHOUT Synthetic Opioids other than Methadone",
  "Cocaine5",
  "Cocaine AND Any Opioid",
  "Cocaine WITHOUT Any Opioid",
  "Cocaine AND Synthetic Opioids other than Methadone",
  "Cocaine WITHOUT Synthetic Opioids other than Methadone",
  "Psychostimulants With Abuse Potential (primarily methamphetamine)6",
  "Psychostimulants With Abuse Potential AND Any Opioid",
  "Psychostimulants With Abuse Potential WITHOUT Any Opioid",
  "Psychostimulants With Abuse Potential AND Synthetic Opioids other than           Methadone",
  "Psychostimulants With Abuse Potential WITHOUT Synthetic Opioids other than Methadone",
  "Benzodiazepines7",
  "Benzodiazepines AND Any Opioid",
  "Benzodiazepines WITHOUT Any Opioid",
  "Benzodiazepines AND Synthetic Opioids other than Methadone",
  "Benzodiazepines WITHOUT Synthetic Opioids other than Methadone",
  "Antidepressants8",
  "Antidepressants AND Any Opioid",
  "Antidepressants WITHOUT Any Opioid",
  "Antidepressants AND Synthetic Opioids other than Methadone",
  "Antidepressants WITHOUT Synthetic Opioids other than Methadone"
  )

list = rep(list, each = 46)
list = tibble(Category = list)


odsGender = od2 |>  pivot_longer(c(2:25), names_to = "year", values_to = "deaths") |>
  filter(!is.na(gender)) |> filter(gender == "Female" | gender == "Male") |> filter(year != "2015-2021 Fold Change") |> select(-Category)

odsGender = tibble(list, odsGender)

```


## Death from Overdose in the United States by Year   


```{r overall deaths from overdoses by sex}
totalods = od2 |> slice(2:3) |>  pivot_longer(c(2:25), names_to = "year", values_to = "deaths") 

totalods$year = as.numeric(totalods$year)
totalods$Category = as.factor(totalods$Category)

a = totalods |> ggplot(aes(x = year, y = deaths, col = Category)) + 
  geom_point(size = 3) +
  geom_line(linewidth = 1) +
  theme_minimal() +
  labs(y = "Deaths from Overdose in the US", x = "Year") +
  scale_color_brewer(palette = "Pastel1")+
  theme(
    axis.title.y = element_text(color = "black", family = "Avenir", size = 20),
    axis.title.x = element_blank(), 
    axis.text.x = element_text(color = "black", size = 20, family = "Avenir"),
    axis.text.y = element_text(color = "black", size = 18, family = "Avenir"),
    legend.title = element_blank(),
    legend.text = element_text(color = "black", family = "Avenir", size = 15),
    legend.position = "top"
  )
```


```{r overall deaths from overdoses by sex figure, fig.height=6, fig.width=12, fig.align='center'}
grid.arrange(a)
```



### Death from Overdose by Substance Type    

#### Opioids  

* 



```{r death from opioids}
totalsODs$year = as.numeric(totalsODs$year)
totalsODs$Category = as.factor(totalsODs$Category)

b = totalsODs |> filter(Category == "Any Opioid1") |>
  ggplot(aes(x = year, y = deaths)) + 
  geom_point(size = 3) +
  geom_line(linewidth = 1) +
  theme_minimal() +
  labs(y = "Overdose Deaths in the US", x = "Year") +
  scale_color_brewer(palette = "Pastel1")+
  theme(
    axis.title.y = element_text(color = "black", family = "Avenir", size = 20),
    axis.title.x = element_blank(), 
    axis.text.x = element_text(color = "black", size = 20, family = "Avenir"),
    axis.text.y = element_text(color = "black", size = 18, family = "Avenir"),
    legend.title = element_blank(),
    legend.text = element_text(color = "black", family = "Avenir", size = 15),
    legend.position = "top"
  )


## gender differences:  
odsGender$year = as.numeric(odsGender$year)
odsGender$Category = as.factor(odsGender$Category)


c = odsGender|> filter(
                     Category == "Prescription Opioids2")  |>  
  ggplot(aes(x = year, y = deaths, col = gender)) +
    geom_point(size = 3) +
  geom_line(linewidth = 1, aes(group = gender)) +
  theme_minimal() +
  labs(y = "Deaths from Overdose in the US", x = "Year", title = "Prescription Opioids", caption = "") +
  scale_color_brewer(palette = "Pastel1") +
  theme(
    axis.title.y = element_text(color = "black", family = "Avenir", size = 20, hjust = 1.1),
    axis.title.x = element_blank(), 
    axis.text.x = element_text(color = "black", size = 20, family = "Avenir"),
    axis.text.y = element_text(color = "black", size = 18, family = "Avenir"),
    legend.title = element_blank(),
    legend.text = element_text(color = "black", family = "Avenir", size = 15),
    legend.position = c(0.8, 1.09),
    legend.direction = "horizontal",
    legend.box = "horizontal",
    strip.text = element_text(color = "black", family = "Avenir", size = 15),
    plot.title = element_text(color = "black", family = "Avenir", size = 20),
    axis.line = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    plot.caption = element_text(color = "black", size = 11, family = "Avenir")
    
  ) 


d = odsGender|> filter(
                     Category == "Prescription Opioids AND Synthetic Opioids other than Methadone")  |>  
  ggplot(aes(x = year, y = deaths, col = gender)) +
    geom_point(size = 3, show.legend = FALSE) +
  geom_line(linewidth = 1, aes(group = gender), show.legend = FALSE) +
  theme_minimal() +
  labs(y = "Deaths from Overdose in the US", x = "Year", title = "Prescription and Synthetic Opioids", caption = "Data shown does not include methadone.") +
  scale_color_brewer(palette = "Pastel1") +
  theme(
    axis.title.y = element_blank(),
    axis.title.x = element_blank(), 
    axis.text.x = element_text(color = "black", size = 20, family = "Avenir"),
    axis.text.y = element_text(color = "black", size = 18, family = "Avenir"),
    legend.title = element_blank(),
    legend.text = element_text(color = "black", family = "Avenir", size = 15),
    legend.position = c(0.8, 1.03),
    legend.direction = "horizontal",
    legend.box = "horizontal",
    strip.text = element_text(color = "black", family = "Avenir", size = 15),
    plot.title = element_text(color = "black", family = "Avenir", size = 20),
    axis.line = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    plot.caption = element_text(color = "black", size = 11, family = "Avenir")
  ) 


e = odsGender|> filter(
                     Category == "Prescription Opioids WITHOUT Synthetic Opioids other than Methadone")  |>  
  ggplot(aes(x = year, y = deaths, col = gender)) +
    geom_point(size = 3, show.legend = FALSE) +
  geom_line(linewidth = 1, aes(group = gender), show.legend = FALSE) +
  theme_minimal() +
  labs(y = " ", x = "Year", title = "Prescription Opioids without Synthetic Opioids", caption = "Data shown does not include methadone.") +
  scale_color_brewer(palette = "Pastel1") +
  theme(
    axis.title.y = element_text(color = "black", family = "Avenir", size = 20),
    axis.title.x = element_blank(), 
    axis.text.x = element_text(color = "black", size = 20, family = "Avenir"),
    axis.text.y = element_text(color = "black", size = 18, family = "Avenir"),
    legend.title = element_blank(),
    legend.text = element_text(color = "black", family = "Avenir", size = 15),
    legend.position = c(0.8, 1.0),
    legend.direction = "horizontal",
    legend.box = "horizontal",
    strip.text = element_text(color = "black", family = "Avenir", size = 15),
    plot.title = element_text(color = "black", family = "Avenir", size = 20),
    axis.line = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    plot.caption = element_text(color = "black", size = 11, family = "Avenir")
  ) 

```


```{r, fig.width=12, fig.height=6, fig.align='center'}
grid.arrange(b)
```



```{r, fig.height=6, fig.width=12}
grid.arrange(c, d, e, nrow = 2)
```



#### Opiates  

*

```{r}

f = odsGender |> filter(Category == "Heroin4") |>
  ggplot(aes(x = year, y = deaths, col = gender)) +
    geom_point(size = 3) +
  geom_line(linewidth = 1, aes(group = gender)) +
  theme_minimal() +
  labs(y = "Deaths from Overdose in the US", x = "Year", title = "Heroin", caption = "") +
  scale_color_brewer(palette = "Pastel1") +
  theme(
    axis.title.y = element_text(color = "black", family = "Avenir", size = 20, hjust = 1.1),
    axis.title.x = element_blank(), 
    axis.text.x = element_text(color = "black", size = 20, family = "Avenir"),
    axis.text.y = element_text(color = "black", size = 18, family = "Avenir"),
    legend.title = element_blank(),
    legend.text = element_text(color = "black", family = "Avenir", size = 15),
    legend.position = c(0.8, 1.05),
    legend.direction = "horizontal",
    legend.box = "horizontal",
    strip.text = element_text(color = "black", family = "Avenir", size = 15),
    plot.title = element_text(color = "black", family = "Avenir", size = 20),
    axis.line = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    plot.caption = element_text(color = "black", size = 11, family = "Avenir")
  ) 

g = odsGender |> filter(Category == "Heroin WITHOUT Synthetic Opioids other than Methadone") |>
  ggplot(aes(x = year, y = deaths, col = gender)) +
    geom_point(size = 3) +
  geom_line(linewidth = 1, aes(group = gender)) +
  theme_minimal() +
  labs(y = "Deaths from Overdose in the US", x = "Year", title = "Heroin without Synthetic Opioids", caption = "") +
  scale_color_brewer(palette = "Pastel1") +
  theme(
    axis.title.y = element_text(color = "black", family = "Avenir", size = 20, hjust = 1.1),
    axis.title.x = element_blank(), 
    axis.text.x = element_text(color = "black", size = 20, family = "Avenir"),
    axis.text.y = element_text(color = "black", size = 18, family = "Avenir"),
    legend.title = element_blank(),
    legend.text = element_text(color = "black", family = "Avenir", size = 15),
    legend.position = c(0.8, 1.09),
    legend.direction = "horizontal",
    legend.box = "horizontal",
    strip.text = element_text(color = "black", family = "Avenir", size = 15),
    plot.title = element_text(color = "black", family = "Avenir", size = 20),
    axis.line = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    plot.caption = element_text(color = "black", size = 11, family = "Avenir")
  )
```


```{r, fig.height=6, fig.width=12, fig.align='center'}
grid.arrange(f, g, nrow = 1)
```





#### Benzodiazapine  






#### Antidepressants  




```{r}
totalsODs |> filter(Category == "Antidepressants8") |>
  ggplot(aes(x = year, y = deaths)) + 
  geom_point(size = 3) +
  geom_line(linewidth = 1) +
  theme_minimal() +
  labs(y = "Overdose Deaths in the US", x = "Year") +
  scale_color_brewer(palette = "Pastel1")+
  theme(
    axis.title.y = element_text(color = "black", family = "Avenir", size = 20),
    axis.title.x = element_blank(), 
    axis.text.x = element_text(color = "black", size = 20, family = "Avenir"),
    axis.text.y = element_text(color = "black", size = 18, family = "Avenir"),
    legend.title = element_blank(),
    legend.text = element_text(color = "black", family = "Avenir", size = 15),
    legend.position = "top"
  )




odsGender|> filter(
                     Category == "Antidepressants WITHOUT Any Opioid" |
                       Category == "Antidepressants8" 
                       ) |>
  ggplot(aes(x = year, y = deaths, col = gender)) +
    geom_point(size = 3) +
  geom_line(linewidth = 1, aes(group = gender)) +
  theme_minimal() +
  labs(y = "Deaths from Overdose in the US", x = "Year") +
  scale_color_brewer(palette = "Pastel1") +
  theme(
    axis.title.y = element_text(color = "black", family = "Avenir", size = 20),
    axis.title.x = element_blank(), 
    axis.text.x = element_text(color = "black", size = 20, family = "Avenir"),
    axis.text.y = element_text(color = "black", size = 18, family = "Avenir"),
    legend.title = element_blank(),
    legend.text = element_text(color = "black", family = "Avenir", size = 15),
    legend.position = "top",
    strip.text = element_text(color = "black", family = "Avenir", size = 15)
  ) +
  facet_wrap(~Category)



```



### Death from Overdose by Polysubstance Use  














Categories:  

Opioids:  


* Any Opioid 
* Select Opioid and synthetic opioids other than methadone  
* Prescription opioids without synthetic opioids other than methadone  
* Synthetic opioids other than methadone  
* Heroin  
* Heroin and synthetic opioids other than methadone  
* Heroin without synthetic opioids other than methdone  
* Stimulants  
* Stimulants and any opioid  
* Stimulants without any opioid  
* Stimulants and Synthetic opioids other than methadone  
* Stimulants without synthetic opioids other than methadone  
* Cocaine  
* Cocaine and any opioid  
* Cocaine without any opioid  
* Cocaine and synthetic opioids other than methadone  
* Cocaine without synthetic opioids other than methadone  
* Psychostimulants with abuse potential (primarily methamphetamine)  
* Psychostimulants with abuse potential and any opioids  
* Psychostimulants with abuse potential without opioids 
* Psychostimulants with abuse potential and synthetic opioids other than methadone  
* Psychostimulants with abuse potential without synthetic opiods other than methadone  
* Benzodiazepines  
* Benzodiazepines and any opioid  
* Benzodiazepines without any opioid  
* Benzodiazepines and synthetic opioids other than methadone  
* Benzodiazepines without synthetic opioids other than methadone  
* Antidepressants and any opioid  
* Antidepressants without any opioid  
* Antidepressants and synthetic opioids other than methadone  
* Antidepressants without synthetic opioids other than methadone 



Opioids: 
* Select Opioid and synthetic opioids other than methadone  
* Prescription opioids without synthetic opioids other than methadone  
* Synthetic opioids other than methadone  











Opiates:  
* Heroin and synthetic opioids other than methadone  
* Heroin without synthetic opioids other than methdone 

Stimulants:  
* Stimulants  
* Cocaine  
* Psychostimulants with abuse potential (primarily methamphetamine)  








Polysubstance use: 

Opiates and Opioids:  
* Heroin and synthetic opioids other than methadone  
* Heroin without synthetic opioids other than methdone  
* Select Opioid and synthetic opioids other than methadone 



















